---
title: "Homework 02: PCA"
author: "Lena Chan, Gaby Branin, Harry Hegeman, Lauren Salzman"
date: "2/10/2020"
output:
  word_document: default
  html_document: default
  pdf_document: default
subtitle: S&DS 363
---
_______________________________________________________________________________


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

cbb <- read.csv("/Users/lenachan/Desktop/S&DS 363/cbb.csv")
cbb

#isolate continuous variables

cbb2 <- cbb[ ,c("G", "W", "ADJOE", "ADJDE", "BARTHAG", "EFG_O", "EFG_D", "TOR", "TORD", "ORB", "DRB", "FTR", "FTRD", "X2P_O", "X2P_D", "X3P_O", "X3P_D", "ADJ_T", "WAB", "SEED", "YEAR")]

#only select variables with complete data

cbb2 <- cbb2[complete.cases(cbb2),]

#compute the correlation matrix

cor(cbb2)
library(corrplot)
corrplot(cor(cbb2), method = "ellipse")

#because they are highly correlated with each other, we will focus on the variables G, W, ADJOE, ADJDE, WAB, BARTHAG, X2P_D, X2P_O, EFG_O, and EFG_D

#subset the data

cbb3 <- cbb2[ ,c("G", "W", "ADJOE", "ADJDE", "WAB", "BARTHAG", "X2P_D", "X2P_O", "EFG_O", "EFG_D") ]

#G: Number of games played

#W: Number of games won

#ADJOE: Adjusted Offensive Efficiency (An estimate of the offensive efficiency (points scored per 100 possessions) a team would have against the average Division I defense)

#ADJDE: Adjusted Defensive Efficiency (An estimate of the defensive efficiency (points allowed per 100 possessions) a team would have against the average Division I offense)

#WAB: Wins Above Bubble (The bubble refers to the cut off between making the NCAA March Madness Tournament and not making it)

#BARTHAG: Power Rating (Chance of beating an average Division I team)

#X2P_O: Two-Point Shooting Percentage

#X2P_D: Two-Point Shooting Percentage Allowed

#EFG_O: Effective Field Goal Percentage Shot

#EFG_D: Effective Field Goal Percentage Allowed

#1. Assess multivariate normality.

names(cbb3[1])

for (i in 1:10){
   qqnorm(cbb3[,i], xlab=names(cbb3[i])) 
}




